Main Analysis

### 2024-05-15 Manuscript Analysis/main_analysis.Rmd
draft of pupil dilation analysis

@@

pupil_roi <- pupil_df |>
  # filtering out timebins below 500 (i.e., keeping timebins above 500 ms)
  dplyr::filter(time_norm >= 500) |>
  dplyr::group_by(subject, trial) |>
  # filtering out timebins 1000 ms after phrase offset
  dplyr::filter(time_norm<= max(time_norm) - 2000) |>
  ungroup() |>
  dplyr::mutate(speaker = factor(speaker, levels = c("Control", "ALS")),
                rep_acc = factor(rep_acc, levels = c("accurate", "inaccurate")),
                phrase = as.factor(code),
                listener = as.factor(subject))

m0_pupil <- bam(normed_pupil ~ speaker +
                  s(time_norm, by = speaker, k = 20) +
                  s(time_norm, by = rep_acc, k = 20) +
                  # Random Smooths for listener and phrase
                  s(time_norm, listener, bs = 'fs', m =1) +
                  s(time_norm, phrase, bs = 'fs', m = 1),
                discrete = T,
                data = pupil_roi)

plot_smooth(m0_pupil,
            view = "time_norm",
            plot_all = c("speaker", "rep_acc"),
            rug = F, se = 1, rm.ranef = T)

```
#### Model Check

@@

>>>>

Initial model with smooths for speaker and phrase.

@@

pupil_roi <- pupil_df |>
  # filtering out timebins below 500 (i.e., keeping timebins above 500 ms)
  dplyr::filter(time_norm >= 500) |>
  dplyr::group_by(subject, trial) |>
  # filtering out timebins 1000 ms after phrase offset
  dplyr::filter(time_norm<= max(time_norm) - 2000) |>
  ungroup() |>
  dplyr::mutate(speaker = factor(speaker, levels = c("Control", "ALS")),
                rep_acc = factor(rep_acc, levels = c("accurate", "inaccurate")),
                phrase = as.factor(code),
                listener = as.factor(subject),
                condition = case_when(speaker == "Control" & rep_acc == "accurate" ~ "Control Accurate",
                                      speaker == "Control" & rep_acc == "inaccurate" ~ "Control Inaccurate",
                                      speaker == "ALS" & rep_acc == "accurate" ~ "ALS Accurate",
                                      TRUE ~ "ALS Inaccurate"),
                condition = factor(condition, 
                                   levels = c("Control Accurate", "Control Inaccurate", "ALS Accurate", "ALS Inaccurate")))

m0_pupil <- bam(normed_pupil ~ condition +
                  s(time_norm, by = condition, k = 20) +
                  # Random Smooths for listener and phrase
                  s(time_norm, listener, bs = 'fs', m =1) +
                  s(time_norm, phrase, bs = 'fs', m = 1),
                discrete = T,
                data = pupil_roi)

summary(m0_pupil)

plot_smooth(m0_pupil,
            view = "time_norm",
            plot_all = "condition",
            rug = F, se = 1, rm.ranef = T)

#### Model Check

@@

### Model 1

Adding in AR(1) structure to address autocorrelation concerns. Also increased knots based on output from the initial model, and took away the random smooth for phrase due to model convergence issues.

# Marking start of trials

pupil_roi <- pupil_roi |>
  dplyr::group_by(listener, speaker) |>
  dplyr::mutate(num_points = n()) |>
  dplyr::mutate(start_event = c(TRUE, rep(FALSE, each = (num_points - 1))))

m1_pupil <- bam(normed_pupil ~ condition +
                  s(time_norm, by = condition, k = 25) +
                  # Random Smooths for listener and phrase
                  s(time_norm, listener, bs = 'fs', m =1),
                discrete = T,
                AR.start = start_event,
                rho = start_value_rho(m0_pupil),
                data = pupil_roi)

summary(m1_pupil)

plot_smooth(m1_pupil,
            view = "time_norm",
            plot_all = "condition",
            rug = F, se = 1, rm.ranef = T)

#### Model Check

Autocorrelation improved. This model's residuals also look better, but there are still some heavy residual tails.

gam.check(m1_pupil)

acf_resid(m1_pupil)

### Model 2

Adding scaled t distribution to fix model fit.

m2_pupil <- bam(normed_pupil ~ condition +
                  s(time_norm, by = condition, k = 25) +
                  # Random Smooths for listener and phrase
                  s(time_norm, listener, bs = 'fs', m =1),
                discrete = T,
                AR.start = start_event,
                rho = start_value_rho(m0_pupil),
                family = "scat",
                data = pupil_roi)

summary(m2_pupil)

m2_prediction <- plot_smooth(m2_pupil,
            view = "time_norm",
            plot_all = "condition",
            rug = F, se = 1, rm.ranef = T)

#### Model Check

Model residuals look better, but not great

gam.check(m2_pupil)

acf_resid(m2_pupil)

#### Difference Plots

plot_diff(m2_pupil, view = "time_norm",
          comp = list(condition = c("ALS Accurate", "Control Accurate")),
          rm.ranef = T)

plot_diff(m2_pupil, view = "time_norm",
          comp = list(condition = c("Control Inaccurate", "Control Accurate")),
          rm.ranef = T)

plot_diff(m2_pupil, view = "time_norm",
          comp = list(condition = c("ALS Inaccurate", "ALS Accurate")),
          rm.ranef = T)

```
## Plot

Draft pupil dilation plot

I am trying to find a way to predict model-predicted values against the actual data. geom_smooth doesn't consider the random effects, so I am thinking I need to extract model data from the final model to show the goodness of fit

It might be good to do this for every iteration of the model to show the process too...

ind_pupil_df <- pupil_df |>
  dplyr::group_by(speaker, subject, rep_acc, time_norm) |>
  summarize(pupil = mean(normed_pupil, na.rm = T),
            se = sd(normed_pupil, na.rm = T)/sqrt(n()))

sum_pupil_df <- ind_pupil_df |>
  dplyr::group_by(speaker, rep_acc, time_norm) |>
  summarize(m_pupil = mean(pupil, na.rm = T),
            se_pupil = sd(pupil, na.rm = T)/sqrt(n()))

# I'm trying to find a way to plot predicted values against actual data.
predicted_values <- m2_prediction[[1]][c("condition", "time_norm", "ul", "ll")]

sum_pupil_df |>
  ggplot() +
  aes(x = time_norm,
      y = m_pupil,
      color = speaker,
      fill = speaker) +
  geom_smooth() +
  geom_line() +
  geom_vline(xintercept = 0, size = 0.35) +
  geom_vline(xintercept = 1910, linetype = 2) +
  annotate("rect", xmin = 500, xmax = 2910, ymin = 0, ymax = 200, alpha = 0.2) +
  coord_cartesian(ylim = c(-10, 250), xlim = c(0, 4910)) +
  facet_wrap("rep_acc") +
  theme_bw() +
  theme(aspect.ratio = 1,
        legend.position = "bottom") +
  labs(x = "Normalized Time (ms)",
       y = "Pupil Dilation (Arbitrary Units)") +
  scale_color_manual(values = c(my_pal[1], my_pal[5])) +
  scale_fill_manual(values = c(my_pal[1], my_pal[5]))

ggsave("Figures/pupil_plot.png", plot = last_plot())
